Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-27 article/main.tex
algo step 1 cas C inconnu

\todo{à fusionnier avec la section suivante!}
Pour casser PCG, nous avons utilisé la méthode présentée par Alan M. Frieze, Johan Hastad, Ravi Kannan, Jeffray C. Lagarias et Adi Shamir \cite{Frieze}. Plaçons nous dans le cas simple du Lehmer's generator\cite{Lehmer}. It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment. The outputs $X$ are the 64 upper-bits of $S$. For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$. 
\todo{rajouter ref Lehmer's generator}

@@

    \State $ S_0[0:64] \gets  S'_0 + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
    \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$

@@

Set $\Delta S_i = S{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). The prediction algorithm proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_4$.
\item Fully reconstruct $\Delta S_0$ from this partial knowledge and $X_0, \dots, X_{39}$.
\item Reconstruct $S_0$ from $\Delta S_0$ and $X_0, \dots, X_{32}$.
\end{enumerate}

@@

In order to access to a part of $\Delta S_i$, we use the same strategy as in
section~\ref{sec:Cknown}: we guess the least significant bits of $S_0$ and some
rotations. The difference is that, in addition, to obtain the ``long-term
advantage'' in the least significant bits of the states, we must also guess the
least significant bits of $c$ as well. We must also guess one additional
rotation to compute the differences $\Delta S_i$.

@@

Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
\[
  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r_i)[0:\ell]~|~r_i=0, 1, \dots, 63 \bigr\}.
\]
If none of the rotations yields a match, then the guesses made beforehand have
to be wrong. As soon as $\ell \geq 6$, this test is efficient (and it can be
done for any number of fresh $X_i$'s). As a consequence, bad guesses can be
filtered deterministically.

\begin{align*}
  \mathbf{S} &= \LCG(S_0, c) \\
             &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
             &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
  \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
             &= \LCG(S_1 - S_0, 0)
\end{align*}

\begin{algorithmic}[1]
  \Procedure{ReconstructPartialDifference}{$X_0, \dots, X_9$}
  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
  \State $T'_j \gets T_j -  S'_i[58:64+\ell]$ \Comment{Shifted truncated LCG}
  \State $\Delta T'_j = T'_{j+1} - T'_j$  \Comment{Difference}
  \State $(\Delta U_0, \Delta U_1, \Delta U_2, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \Delta T'_1, \Delta T'_2, \Delta T'_3) \cdot 2^{39} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot U_0[0:58-\ell]$
  \State ...
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}

\begin{tikzpicture}[yscale=0.66]
  \draw[red, use as bounding box] (-0.75, -3.75) rectangle (12.8, 2.25);

  % S_i
  \begin{scope}
    % remplissage
    \fill[fill=lightgray] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=lightgray] (0, 0) rectangle +(0.6, 1);
    \fill[fill=lightgray] (11, 0) rectangle +(1.8, 1);

    % déco autour
    \node at (-0.5, 0.5) {$S_i$};
    \draw[<->] (0, 1.5) -- node[above] {64 bits} +(6.375, 0);
    \draw[<->] (6.425, 1.5) -- node[above] {64 bits} +(6.375, 0);
  \end{scope}

  % T_i
  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {00000000 \dots 00000000} (6.4, 1);

    % déco
    \node at (-0.5, 0.5) {$T_i$};
    \draw[<->] (0, -0.5) -- node[below] {$\ell$ bits} +(1.775, 0);
    \draw[<->] (1.8, -0.5) -- node[below] {6} +(0.6, 0);
    \draw[<->] (0, -1.5) -- node[below] {64 bits} +(6.375, 0);
  \end{scope}

\end{tikzpicture}

\subsection{Part 1 : Premier filtre}

Pour tous les $r_i$, les $S_0[0 : \ell]$ et les $C[0 : \ell]$ possibles:
\begin{enumerate}
  \item On récupère le $T$  correspondant à nos suppositions. On calcule $\Delta T$ :
  \begin{equation}
    \Delta T_i = (S_{i+1}[58 : 64 + \ell] - S_i[58 : 64 + \ell] - (S_0[0:\ell] * (A^{i+1} - A^i) + C[0:\ell] * A^i )[58 : 64 + \ell]) * 2^{64 - l - 6}
  \end{equation}
  Ou plus simplement $\Delta T_i = T'_{i+1} - T'_{i}$ avec $T'$ dont on aurait soustrait la composante en $C[0:\ell]$ et non en $C$.

  \item We get the correspondant $\Delta S_0S_1[0 : 64 + \ell]$ by using the babai rounding technique on $\Delta T$ using $G$ and $G^{-1}$.

  \item On teste s'il existe pour l'itération suivante au moins une rotation $r_n$ cohérente avec le $\Delta T$ trouvé. Pratiquement, pour chacune des 64 $r_n$ possibles, on teste si :
  \begin{align}
     \Delta S_0S_n[0:l] &= S_0[64:64 + \ell] \oplus \mathtt{ROT(X_n , 64 - r_n)[0:\ell]}\\
     r_n &= S_0[58:64] \oplus \mathtt{ROT(X_n , 64 - r_n)[58:64]}
  \end{align}
  Si un de ces égalité n'est pas vérifiée, la solution n'est pas retenue, on passe à la prochaine itération.

  \item On répète la dernière opération encore 3 fois.
\end{enumerate}
Cette première partie de l'algorithme comporte l'énorme majorité du temps de calcul, c'est donc sur elle que doit se porter la recherche de performance.

>>>>

\todo{à fusionnier avec la section suivante!}
Pour casser PCG, nous avons utilisé la méthode présentée par Alan M. Frieze, Johan Hastad, Ravi Kannan, Jeffray C. Lagarias et Adi Shamir \cite{Frieze}. Plaçons nous dans le cas simple du Lehmer's generator \todo{FIXME}. It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment. The outputs $X$ are the 64 upper-bits of $S$. For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$. 
\todo{rajouter ref Lehmer's generator}

@@

    \State $ S_0[0:64] \gets  S'_0[0:64] + 2^{\ell} \cdot U_0[0:64-\ell]$ \Comment{Reconstruct $S_0$}
    \State $ S_0[64:128] \gets S_0[0:64] \oplus Y_0$

@@

Set $\Delta S_i = S_{i+1} - S_i$; it is easily checked that $\Delta S_i$ is a
geometric progression of common ratio $a$. Therefore, reconstructing both $S_0$
and $\Delta S_0$ is sufficient to compute all subsequent states (and recover the
unknown increment $c$). We note that
\begin{align*}
  S_i - S_0 \equiv \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{align*}

The prediction algorithm proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{39}$.
\item Fully reconstruct $\Delta S_0$ from this partial knowledge and the outputs.
\item Reconstruct $S_0$ from $\Delta S_0$ and the outputs.
\end{enumerate}

@@

In order to access to a part of $\Delta S_i$, we use the same strategy as in
section~\ref{sec:Cknown}: we guess the least significant bits of $S_0$ and some
rotations. The difference is that, in addition, to obtain the ``long-term
advantage'' in the least significant bits of the states, we must also guess the
least significant bits of $c$ as well ($c$ is always odd; this makes one less
bit to guess). We must also guess one additional rotation to compute the
differences $\Delta S_i$.

@@

Given a ``fresh'' output $X_i$, and assuming that the guesses are correct, then we should have:
\begin{equation}\label{eq:consistency}
  S_i[0:\ell] \oplus S_i[64:64+\ell] \in \bigl\{ (X_i \lll r_i)[0:\ell]~|~r_i=0, 1, \dots, 63 \bigr\}.
\end{equation}
If none of the rotations yields a match, then the guesses made beforehand have
to be wrong. As soon as $\ell \geq 6$, this test is efficient (and it can be
done for any number of fresh $X_i$'s). As a consequence, bad guesses can be
filtered deterministically.

\begin{algorithmic}[1]
  \Procedure{ConsistencyCheck}{$\Delta S_0, w_0, Y_0, X_5, \dots, X_k$}
  \State // \emph{Implements the test given by eq.~\eqref{eq:consistency}}.
  \State $x_0 = w_0 \oplus Y_0[0:\ell]$ \Comment{$x_0 = S_0[64:64+\ell]$}
  \For{$i=5, \dots, k$}
  \State $x_i = x_0 + \left(\Delta S_0 \frac{a^i-1}{a-1} \bmod 2^{64+\ell}\right)[64:64+\ell]$. \Comment{$x_i = S_i[64:64+\ell]$}
%  \State $\Delta S_i[\ell:64+\ell] \gets a^i \Delta U_0 \bmod 2^{64}$
  \State $\mathrm{Target} \gets S_i[0:\ell] \oplus x_i$
  \If{$\mathrm{Target} \not\in \{ (X_i \lll r_i)[0:\ell]~|~ 0\leq r_i < 64\}$}
  % \State $\mathrm{Found} \gets \mathbf{False}$
  % \For{$0 \leq r_i < 64$}
  % \State $Y_i \gets X_i \lll r_i$
  % \If{$Y_i[0:\ell] = \mathrm{Target}$}
  % \State $\mathrm{Found} \gets \mathbf{True}$ 
  % \EndIf
  % \EndFor
  % \If{not Found}
  \State \Return \textbf{False} \Comment{Bad Guesses}
  \EndIf
  \EndFor
  \State \Return \textbf{True} \Comment{Good  Guesses}
  \EndProcedure

\State

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
  \State $S'_j \gets a^j w_0 + (2c+1) \sum_{k = 0}^j a^k \bmod 2^{128}$ \Comment{Shifted states}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(w_j \oplus Y_j[0:\ell]\right)$ \Comment{Truncated LCG}
  \State $T'_j \gets (T_j -  S'_i[58:64+\ell]) \bmod 2^{6+\ell}$ \Comment{Shifted truncated LCG}
  \State $\Delta T'_j = (T'_{j+1} - T'_j) \bmod 2^{6 + \ell}$  \Comment{Difference}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde G_4^{-1} \right\rceil \cdot \widetilde G_4$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left((w_1 - w_0) \bmod 2^{\ell}\right) + 2^{\ell} \cdot \Delta U_0[0:58-\ell]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}

\begin{theorem}
  If $\ell \geq 13$, then algorithm \textsc{ReconstructPartialDifference} is
  correct (it always returns the correct values).
\end{theorem}

\begin{proof}
  \todo{WRITE-ME}
  \begin{align*}
  \mathbf{S} &= \LCG(S_0, c) \\
             &= \LCG\left(S_0[0:\ell] + 2^\ell S_0[\ell:128], c[0:\ell] + 2^\ell c[\ell:128]\right) \\
             &= \LCG\left(S_0[0:\ell], c[0:\ell]\right) + 2^\ell \LCG\left(S_0[\ell:128],  c[\ell:128]\right) \\
  \Delta \mathbf{S} &= \LCG(S_1, c) - \LCG(S_0, c) \\
             &= \LCG(S_1 - S_0, 0)
\end{align*}
\end{proof}
